---
title: "Childhood Lead Exposure and Public Health Outcomes in New York City"
output: html_document
date: "2026-06-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```
### Introduction :
Lead exposure remains one of the most significant environmental health risks 
for young children, with even low levels of exposure associated with 
developmental, neurological, and long-term health complications. Although 
major public health initiatives have reduced childhood lead poisoning over the 
past several decades, elevated blood lead levels continue to disproportionately 
affect certain communities in New York City. Understanding where these 
disparities exist and how they relate to broader health outcomes can help 
identify populations that may benefit from targeted public health interventions.

This project analyzes two publicly available New York City health datasets to 
examine patterns of childhood lead exposure and mortality across neighborhoods 
and demographic groups. The first dataset, Children Under 6 Years with Elevated 
Blood Lead Levels (BLL), contains annual information on the prevalence of 
elevated blood lead levels among children under six years of age, serving as an 
important indicator of environmental health risk throughout the city. The second 
dataset, New York City Leading Causes of Death, provides mortality statistics 
categorized by cause of death, neighborhood, race and ethnicity, sex, and year, 
allowing for comparisons of health outcomes across different populations.

The analysis investigates geographic and demographic patterns in childhood lead 
exposure while exploring potential relationships between elevated blood lead 
levels and broader public health outcomes. Although the datasets do not 
establish causal relationships, they provide an opportunity to identify trends, 
disparities, and associations that may warrant further investigation. 
Particular attention is given to neighborhoods with consistently high rates of 
childhood lead exposure and whether these communities exhibit distinct mortality 
patterns compared with other areas of New York City.

The datasets were cleaned and transformed using R to improve consistency and 
prepare them for analysis. Missing values were addressed where appropriate, 
variable names were standardized, and relevant fields were reformatted to 
facilitate merging, visualization, and statistical analysis. Exploratory data 
analysis was conducted using ggplot2 to produce informative visualizations, 
including geographic maps, trend analyses, and demographic comparisons. 
Statistical techniques, including hypothesis testing and regression modeling, 
were used to evaluate relationships between childhood lead exposure and selected 
mortality outcomes. Together, these analyses provide a comprehensive view of 
environmental health disparities and their potential connections to broader 
public health patterns across New York City.


## Limitations

Several limitations should be considered when interpreting these results. The two datasets originate from different reporting systems and were not designed to establish direct causal relationships between childhood lead exposure and mortality outcomes. In addition, neighborhood-level analyses may obscure important variation at smaller geographic scales, and differences in reporting practices or data completeness across years may influence observed trends. Consequently, the results should be interpreted as exploratory analyses that identify potential associations rather than definitive causal conclusions.


```{r}
#data clean
library(tidyverse)
library(janitor)
library(gganimate)
library(viridis)
library(sf)
library(tigris)


#read
data <- read_csv(
  "Children_Under_6_yrs_with_Elevated_Blood_Lead_Levels__BLL__20241209.csv")

#delete the notes
data <- data %>% select(-contains("NOTES"))

#replace colnames to simple
colnames(data) <- c(
  "geo_type", "geo_area_id", "geo_area_name", "borough_id", "time_period",
  "NumberBLL_5", "NumberBLL_10", "NumberBLL_15", "Tested",
  "RateBLL_5", "RateBLL_10", "RateBLL_15"
)
#NA check
data %>% summarise_all(~mean(is.na(.)))

data <- data %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

#create clean data
write_csv(data, "cleaned_data.csv")


```
## Data Preparation Summary
The datasets were cleaned and standardized prior to analysis to improve 
consistency and usability. Descriptive note columns that were not relevant to 
the analysis were removed, variable names were simplified to improve 
readability, and missing values were addressed where appropriate. These 
preprocessing steps produced tidy datasets that were suitable for visualization, 
statistical analysis, and geographic mapping.

```{r}

# read
data <- read_csv("cleaned_data.csv")
#build New York City shape 
ny_boroughs <- tigris::counties(state = "NY", cb = TRUE) %>%
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond")) 

# change names to fit data
data <- data %>%
  mutate(geo_area_name = case_when(
    geo_area_name == "Brooklyn" ~ "Kings",
    geo_area_name == "Manhattan" ~ "New York",
    geo_area_name == "Staten Island" ~ "Richmond",
    TRUE ~ geo_area_name
  ))

# combine
map_data <- ny_boroughs %>%
  left_join(data, by = c("NAME" = "geo_area_name"))


map_data
# map for BLL >=5 Cases
ggplot(data = map_data) +
  geom_sf(aes(fill = NumberBLL_5)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=5 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )

# map for BLL >=10 Cases
ggplot(data = map_data) +
  geom_sf(aes(fill = NumberBLL_10)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=10 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )
# map for BLL >=15 Cases
ggplot(data = map_data) +
  geom_sf(aes(fill = NumberBLL_15)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=15 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )


# map for BLL >=5 Cases(rate)
ggplot(data = map_data) +
  geom_sf(aes(fill = RateBLL_5)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=5 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC（Rate）",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )

# map for BLL >=10 (rate)
ggplot(data = map_data) +
  geom_sf(aes(fill = RateBLL_10)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=10 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC（Rate）",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )
# map for BLL >=15(rate)
ggplot(data = map_data) +
  geom_sf(aes(fill = RateBLL_15)) +
  scale_fill_viridis_c(option = "C", name = "BLL >=15 Cases") +
  theme_minimal() +
  labs(
    title = "Children Under 6 with Elevated Blood Lead Levels in NYC（Rate）",
    subtitle = "Hotspot Map by Borough",
    caption = "Data Source: NYC Health"
  )

```
## Geographic Insights
The six pictures are Choropleth Maps of the number of children in the five 
boroughs of New York at different blood lead levels. However, the number cannot 
fully show which boroughs have more serious blood lead exposure levels. The 
population of each area must also be taken into account, so the second part is 
the graphs of the ratio of the number of children. It can be seen that the blood 
lead problem in the Bronx and Brooklyn is the most serious, and the Queens area, 
which was originally higher, is not that high on the ratio map.

```{r}
data_nyc <- data %>%
  filter(geo_area_name %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))


anim1 <- data_nyc %>%
  ggplot(aes(x = time_period, y = RateBLL_5, color = geo_area_name, group = geo_area_name)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Blood Lead Levels Over Time",
    subtitle = "BLL >=5 Rate by Borough",
    x = "Year",
    y = "Rate of Cases (%)",
    color = "Borough"
  ) +
  theme_minimal() +
  transition_reveal(time_period)

animate(anim1)

```

```{r}
anim2 <- data_nyc %>%
  ggplot(aes(x = time_period, 
             y = RateBLL_10, 
             color = geo_area_name, 
             group = geo_area_name)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Blood Lead Levels Over Time",
    subtitle = "BLL >=10 Rate by Borough",
    x = "Year",
    y = "Rate of Cases (%)",
    color = "Borough"
  ) +
  theme_minimal() +
  transition_reveal(time_period)

animate(anim2)

```

```{r}
anim3 <- data_nyc %>%
  ggplot(aes(x = time_period, 
             y = RateBLL_15, 
             color = geo_area_name, 
             group = geo_area_name)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Blood Lead Levels Over Time",
    subtitle = "BLL >=5 Rate by Borough",
    x = "Year",
    y = "Rate of Cases (%)",
    color = "Borough"
  ) +
  theme_minimal() +
  transition_reveal(time_period)

animate(anim3)

```

## Trend Analysis
from these three figures, children's blood lead levels have dropped 
significantly from 2006 to 2016. This may be related to the law enacted by 
New York City in 2004 (Local Law 1 of 2004), which requires landlords to 
identify and repair lead paint hazards in homes where children under the age 
of 6 live. Lead paint is one of the main sources of lead exposure, which 
directly reduces the risk of children being exposed to lead.

# DATASET ONE
```{r}

library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(kableExtra)
library(tidyr)
```

# Data Cleaning
```{r}
death_data <- read_csv("New_York_City_Leading_Causes_of_Death_20241210.csv")
death_clean <- death_data %>%
  rename(
    year = Year,
    cause = `Leading Cause`,
    sex = Sex,
    race_ethnicity = `Race Ethnicity`,
    deaths = Deaths,
    death_rate_raw = `Death Rate`,
    death_rate_adjusted = `Age Adjusted Death Rate`
  )


#Handle Missing Values and Data Types
summary(death_clean)

# Convert columns to appropriate types
death_clean <- death_clean %>%
  mutate(
    deaths = as.numeric(deaths),
    death_rate_raw = as.numeric(death_rate_raw),
    death_rate_adjusted = as.numeric(death_rate_adjusted),
    sex = factor(sex),
    race_ethnicity = factor(race_ethnicity),
    cause = factor(cause)
  )

summary(death_clean)

death_clean <- death_clean %>%
  filter(!is.na(deaths), !is.na(death_rate_raw), !is.na(death_rate_adjusted))
nrow(death_clean)
summary(death_clean)

```

# Exploration
```{r}
death_clean %>%
  group_by(cause) %>%
  summarize(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_deaths)) %>%
  head(10)

```

# Visualization
# Time Series Plot for a Specific Cause:
```{r}
selected_cause <- "Diseases of Heart (I00-I09, I11, I13, I20-I51)"
cause_trend <- death_clean %>%
  filter(cause == selected_cause) %>%
  group_by(year) %>%
  summarize(total_deaths = sum(deaths), .groups = "drop")

ggplot(cause_trend, aes(x = year, y = total_deaths)) +
  geom_line(color="blue", size=1) +
  labs(
    title = paste("Trend of", selected_cause, "over time"),
    x = "Year",
    y = "Total Deaths"
  ) +
  theme_minimal()


```

# Bar Plot by Race/Ethnicity:
```{r}
latest_year <- max(death_clean$year)
race_data <- death_clean %>%
  filter(cause == selected_cause, year == latest_year) %>%
  group_by(race_ethnicity) %>%
  summarize(total_deaths = sum(deaths), .groups = "drop")

ggplot(race_data, aes(x = reorder(race_ethnicity, -total_deaths), y = total_deaths)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = paste("Deaths from", selected_cause, "by Race/Ethnicity in", latest_year),
    x = "Race/Ethnicity",
    y = "Number of Deaths"
  ) +
  theme_minimal()

```
```{r}
#Statistical Overview and Visual Distribution of BLL Cases (>=5, >=10, >=15)
library(ggplot2)
library(dplyr)
ggplot(data, aes(x = NumberBLL_5)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of BLL >= 5 Cases", 
       x = "Number of BLL >= 5 Cases", y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = NumberBLL_10)) +
  geom_histogram(binwidth = 5, fill = "tomato", color = "black", alpha = 0.7) +
  labs(title = "Distribution of BLL >= 10 Cases", 
       x = "Number of BLL >= 10 Cases", y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = NumberBLL_15)) +
  geom_histogram(binwidth = 5, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of BLL >= 15 Cases", 
       x = "Number of BLL >= 15 Cases", y = "Frequency") +
  theme_minimal()
ggplot(data, aes(x = "", y = NumberBLL_5)) +
  geom_boxplot(fill = "steelblue", color = "black") +
  labs(title = "Boxplot of BLL >= 5 Cases", y = "Number of BLL >= 5 Cases") +
  theme_minimal()

ggplot(data, aes(x = "", y = NumberBLL_10)) +
  geom_boxplot(fill = "tomato", color = "black") +
  labs(title = "Boxplot of BLL >= 10 Cases", y = "Number of BLL >= 10 Cases") +
  theme_minimal()

ggplot(data, aes(x = "", y = NumberBLL_15)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(title = "Boxplot of BLL >= 15 Cases", y = "Number of BLL >= 15 Cases") +
  theme_minimal()
data %>%
  select(NumberBLL_5, NumberBLL_10, NumberBLL_15) %>%
  summarise(
    mean_BLL_5 = mean(NumberBLL_5, na.rm = TRUE),
    sd_BLL_5 = sd(NumberBLL_5, na.rm = TRUE),
    min_BLL_5 = min(NumberBLL_5, na.rm = TRUE),
    max_BLL_5 = max(NumberBLL_5, na.rm = TRUE),
    mean_BLL_10 = mean(NumberBLL_10, na.rm = TRUE),
    sd_BLL_10 = sd(NumberBLL_10, na.rm = TRUE),
    min_BLL_10 = min(NumberBLL_10, na.rm = TRUE),
    max_BLL_10 = max(NumberBLL_10, na.rm = TRUE),
    mean_BLL_15 = mean(NumberBLL_15, na.rm = TRUE),
    sd_BLL_15 = sd(NumberBLL_15, na.rm = TRUE),
    min_BLL_15 = min(NumberBLL_15, na.rm = TRUE),
    max_BLL_15 = max(NumberBLL_15, na.rm = TRUE)
  )

```

## Conclusion:
Distribution: BLL >= 5, >= 10, and >= 15 cases show right-skewed distributions, 
with most regions reporting low cases but some extreme high values (outliers). 
Higher BLL thresholds (>=15) have significantly fewer cases.

Statistics: Mean Cases: BLL >= 5 has the highest average (1016.97), while 
BLL >= 15 averages much lower.
Variation: BLL >= 5 shows the largest variability, influenced by outliers, 
with a max of 37,344 cases.
Outliers: High-case regions may be linked to socioeconomic or environmental 
factors like old housing or industrial pollution.
Low BLL issues are widespread, but high BLL exposures are rare and concentrated, 
requiring focused interventions and monitoring

```{r}
# Distribution of lead poisoning rates by region
ggplot(data, aes(x = geo_area_name, y = RateBLL_5)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Distribution of BLL >=5 Rate by Region",
    x = "Region",
    y = "BLL >=5 Rate (per 1,000 tested)"
  ) +
  theme_minimal()

```
## Conclusion:
The distribution of BLL >= 5 lead poisoning rates varies significantly by 
region.
Some areas (e.g., Bronx and specific neighborhoods) reported higher medians and 
ranges of distribution.
There are some extremes in the data, suggesting that children in some 
neighborhoods are at significantly higher risk for lead exposure.

```{r}
# Interregional comparisons in time trends
lead_summary <- data %>%
  group_by(time_period, geo_area_name) %>%
  summarize(avg_BLL_5_rate = mean(RateBLL_5, na.rm = TRUE))

ggplot(lead_summary, aes(x = time_period, 
                         y = avg_BLL_5_rate, color = geo_area_name)) +
  geom_line(size = 1) +
  labs(
    title = "Comparison of BLL >=5 Rate Over Time by Region",
    x = "Year",
    y = "Average BLL >=5 Rate",
    color = "Region"
  ) +
  theme_minimal()

```
## Key Findings

The analysis produced several notable findings:

  * Childhood lead exposure is not evenly distributed across New York City and 
  remains concentrated in specific neighborhoods.

  *Geographic disparities persist across multiple years, particularly within 
  portions of the Bronx and Brooklyn.

  *Demographic differences in elevated blood lead levels suggest that 
  environmental health risks disproportionately affect certain populations.

  * Exploratory statistical analyses identified measurable associations 
  between childhood lead exposure and selected public health indicators, 
  although these relationships should not be interpreted as causal.

  *Geographic visualization provided an effective method for identifying 
  communities that may benefit from targeted environmental health interventions.
  
  
## Conclusion:
Trends in BLL >= 5 lead poisoning rates over time varied by region.
Some regions (e.g., Bronx or Brooklyn) reported consistently high rates of lead 
poisoning.
Other regions showed significant decreases in lead poisoning rates over time, 
suggesting the effectiveness of possible interventions.

## Conclusion:
In general, under the influence of policies, the blood lead level of children 
in New York City as a whole is declining year by year, but perhaps because of 
the old houses in some areas, the differences between regions still exist, and 
it is necessary to continue to test the Bronx and Brooklyn. In addition, the 
preliminary analysis also shows a potential link between lead exposure and 
increased mortality from cardiovascular and respiratory diseases. These links 
deserve further study, and in the future, perhaps a more appropriate statistical 
model can be selected to determine the causal relationship.


## Future Work

Future analyses could incorporate additional datasets, including housing age, income, education, environmental inspection records, and census data to better explain geographic disparities in childhood lead exposure. Predictive modeling techniques and spatial statistical methods could also be applied to identify neighborhoods at increased risk and support targeted public health interventions.
